Distribution of dinucleotides in the human and mouse reference genomes

### 2014-11-22 Rmarkdown/appendix.Rmd
Updated appendex

---
title: "Appendix"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

@@

>>>>

---
title: "Appendix"
author: "Peter Hickey"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
    keep_tex: no
  html_document:
    keep_md: yes
bibliography: ../latex/phd_thesis.bib
---

## CpGs are underrepresented in the human genome

This short R script demonstrates that CpGs are underrepresented in the human reference genome (build hg19).

Firstly, the frequency of each nucleotide is computed. Secondly, the expected frequency of each dinucleotide is computed under the _independence model_. Under the independence model, the frequency of each dinucleotide is simply the product of the respective observed nucleotide frequencies. Finally, these expected frequencies are compared to the observed values from the respective reference genomes.

I only use the autosomes, sex chromosomes and mitochondrial chromosome in all analyses, that is, I do not consider the "random", "unplaced" and "alternate haplotype" contigs in the reference genomes.

All results are computed at the chromosome-level as well as being summarised genome-wide.

### Compute the nucleotide and dinucleotide frequencies

```{r, cache = TRUE}
library(BSgenome.Hsapiens.UCSC.hg19)
library(knitr)
# Nucleotide frequencies
nf <- bsapply(new("BSParams", X = Hsapiens, FUN = letterFrequency, 
            exclude = c("rand", "Un", "hap"), simplify = TRUE), 
            letters = c("A", "C", "G", "T"))
# Pretty print the results
kable(data.frame(percentage = round(100 * rowSums(nf) / 
                                      sum(as.numeric(nf)), 1)))

# GC-percentage
round(100 * sum(nf[c("C", "G"), ]) / sum(as.numeric(nf)), 1)

# Dinucleotide frequencies
df <- bsapply(new("BSParams", X = Hsapiens, FUN = dinucleotideFrequency, 
            exclude = c("rand", "Un", "hap"), simplify = TRUE))
# Pretty print the results
kable(data.frame(percentage = round(100 * rowSums(df) / 
                                      sum(as.numeric(df)), 1)))

# CpG percentage
round(100 * rowSums(df) / sum(as.numeric(df)), 1)["CG"]
```

### Plot the observed:expected ratio for each dinucleotide

```{r dinucleotide_distribution, fig.lp = "fig:dinuclotide_distribution"}
library(ggplot2)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)
edf <- nf %*% t(nf) / (sum(nf %*% t(nf))) * 100
x <- data.frame(Dinucleotide = names(rowSums(df)),
                Observed = as.vector(100 * rowSums(df) / sum(as.numeric(df))), 
                Expected = as.vector(edf))
ggplot(x, aes(x = Dinucleotide, y = Observed / Expected)) + 
  geom_point(size = 5) + ylim(c(0, 1.5)) + 
  ggtitle("CpGs are underrepresented in human reference genome (hg19)") + 
  thesis_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  geom_hline(yintercept = 1, linetype = 2)
```

@@

## Software details

```{r}
devtools::session_info()
```

### 2014-11-22 Rmarkdown/dinucleotide_distribution/dinucleotide_distribution.Rmd
Described methylKit and (most of) DSS
Updated appendex

Distribution of dinucleotides in the human and mouse reference genomes
========================================================

@@

Human genome 
========================================================

Step 1: Compute the nucleotide frequencies
--------------------------------------------------------
```{r}
#### Load necessary packages ####
library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg19)

@@

Step 2: Compute expected number of CpGs under an independence model
--------------------------------------------------------

Software details
========================================================

>>>>

---
title: "Distribution of dinucleotides in the human and mouse reference genomes"
author: Peter Hickey
date: 22 April 2014
output: pdf_document
---

## Distribution of dinucleotides in the human and mouse reference genomes

@@

### Human genome

#### Step 1: Compute the nucleotide frequencies
```{r}
#### Load necessary packages ####
library(BSgenome.Hsapiens.UCSC.hg19)

@@

#### Step 2: Compute expected number of CpGs under an independence model

### Software details

```{r}
devtools::session_info()
```
